Thesis: draft

### 2021-05-03 draft.tex
made some small changes in ch. 1 and ch. 2

\renewcommand{\abstractname}{Abstract}
\selectlanguage{english}
\begin{abstract}
\addcontentsline{toc}{chapter}{Abstract}
The past decade, the rise of social media has become undeniable. More and more people are getting engaged on some type of social media platform and hence, it is probably not wrong to say that nowadays most of us have, beside our real life, a virtual life on social media. This rapidly increased engagement on social media undoubtedly has many advantages. However, every upside has a downside and social media is no different. This thesis deals with the impact of social media on the formation and evolution of opinions in society.\\ 
Social media allows for the spread of information at a faster pace and at an unprecedented scale. However, people are still bound by time- and cognitive constraints. In order to handle the information overflow, social media companies design filtering algorithms that can fine-tune and individualize the information someone is exposed to. This filtering and individualizing of information may have some potentially dangerous downsides in the sense that people might get trapped in ``echo chambers'' that confirm their own beliefs. It is important to know the effects of such algorithmic personalization on the appearance and formation of phenomena such as polarization, spread of extremism, formation of echo chambers, etc.\\
People are, however, more than sheep that thoughtlessly follow their shepherd (contrary to what some books might lead us to believe). They are capable of critical thinking and may be hesitant to change their current opinion. This aspect of potential stubbornness should not be ignored.\\
In this thesis toy models are designed to investigate the interplay between network structure, filtering algorithms and individual resistance to change in the evolution of two competing opinions.\\
We will investigate whether network properties such as community structure and clustering coefficient may enhance or hamper the formation of echo chambers. We will use some basic filtering algorithms to determine their possible effects on the appearance of echo chambers and we will investigate the impact of individual stubbornness. \\
Beside that, we will also pay some attention to the possible ways of distributing the opinions and the stubborn actors and how this may effect the opinion dynamics in the system. \\
Finally, we will test the model on real-world networks. Some question that may arise are: Do the results agree with those obtained from the network models? What are the similarities or differences in the results? What are the possible underlying reason/structures for these similarities/differences?\\

@@

The past decade social media has become ever more important in our daily lives. Nowadays most of us have, beside a real-world life, a virtual life on social media. This increased engagement on social media may have a huge impact on the formation of opinions, not only on the individual scale, but also on a broader, societal scale. \\
\newline
It is known that individuals not only form their opinions through self-reflection, but also through interactions with other people and with their surroundings \cite{Perra2019}. The broad range of interactions people undergo on social media with people from all over the world can thus not be underestimated in the process of opinion formation. The understanding of the role of social media on the emergence of polarization and extremism in society is of uttermost importance.\\
\newline
It is also known that some people change their opinions more easily than others, that is, once in a while you encounter somebody that is really stubborn and persistent towards its own opinion. One of the aims of this thesis is to design toy models of opinion dynamics with stubborn actors on theoretical and real-world social networks to analyze the interplay between individual resistance to change and the network structure in the evolution of two competing opinions.\\
\newline
One important feature of opinion dynamics on social media is the use of algorithmic personalization. Social media allow for simultaneous or asynchronous interactions between people without geographical constraints and thus allow for the spread of information at a faster pace and at an unprecedented scale \cite{Perra2019}. People are, however, still bound by temporal and cognitive constraints. In order to assure a more pleasant/convenient experience on their platforms, social media companies use algorithms to order/filter the posts on an individual's time-line according to what might be relevant to that individual \cite{Perra2019}. It is of great importance to know the possible effects of these filtering algorithms on the evolution of opinions in the population.\\  
\newline
Opinion dynamics models have two important layers. On one hand, we introduce social networks to describe the underlying structure of social interactions; on the other hand, we need appropriate opinion dynamics/formation models to reproduce the opinion dynamics and opinion formation processes found in real life.\\
\newline
Section \ref{complNet} gives an introduction to the first layer, namely that of complex networks. The second layer, the opinion dynamics layer, is introduced in Section \ref{Opinion}. Section \ref{statPhys} introduces concepts such as statistical and social physics. Finally, in Section \ref{goal}, the hypotheses and goals of this thesis are summarized and discussed.

@@

Some other important concepts in network theory are the degree distribution, clustering/community structure, connectivity, etc. These concepts will be explained in depth in Chapter \ref{chap2}, Section \ref{netDef}, but let us already anticipate a bit on the case of social networks. It is found that many real-life social systems have a power-law degree distribution \cite{Muchnik2013}. This heterogeneous or scale-free degree distribution represents one of the three general properties of social networks. The other two are short distances and high clustering. These two are also referred to as ``small world phenomenon'' (note that in the literature the term ``small world'' sometimes only refers to the property of short distances, however, in this thesis, we will use it to indicate both a high clustering and short distances) \cite{Muchnik2013}. Ideally, our theoretical network should exhibit these three properties. Some theoretical networks, that are thoroughly studied, do not possess all of these three properties. However, they might still be used, because of their simplicity and ability to produce analytic results. When using these models one must always bear in mind their limitations to reproduce some properties encountered in real social systems. \\

@@

As stated before, the formation of an individual's opinion is the result of the interplay of many factors such as self-reflection, peer pressure, the individual's personality (eg. stubbornness), the information someone is exposed to, etc.  Opinion dynamics models should try to capture these complex processes in a simplified way. Several models have been developed ranging from simple binary models to more complex, continuous approaches \cite{Sirbu2016}.\\
\newline
The basic idea of all opinion dynamics models is that the nodes or agents in a social network have a variable that represents their opinion and that is updated according to some predefined rules. These models are obviously a simplification of real-world opinion dynamics. It is however shown that they do display a lot of aspects of real opinion formation such as agreement, transitions between order (consensus) and disorder (fragmentation), polarization, formation of echo chambers (clusters of people with the same opinion), etc \cite{Sirbu2016}. \\
\newline
The opinion of the agents in the model can be either discrete or continuous. This thesis deals with models where each agent can have one of the two opinions A or B and thus only deals with the case of discrete opinions. This might come across as a huge simplification of the real-world complexity of opinion formation, but also in real life people often have to choose between two competing opinions (eg. republicans or democrats, cat or dog, renting or buying a house,...). \\
\newline
The rules that determine how an agent updates his or her opinion can include many different aspects. They can, for example, be as simple as a majority rule (i.e. the majority model: if the majority of your neighbors have a certain opinion, you adopt that opinion as well) or can include a more probabilistic way of updating. A small overview of possible opinion dynamics models is given in Chapter \ref{chap3}, Section \ref{Bin} and Section \ref{Con}. \\
\newline
One can also include a resistance parameter that determines the hesitation of an agent to change to a new opinion and/or a parameter that determines the influence of an agent on others. Furthermore, the agents can have a parameter that determines whether they are active or not. The introduction of stubborn actors will be discussed in Chapter \ref{chap3}, Section \ref{stubb}; the activation mechanism of nodes is discussed in Chapter \ref{chap3}, Section \ref{actMech}.\\
\newline
Since this thesis deals with the particular case of opinion dynamics on on-line social platforms, it is important to define/set up different filtering algorithms. The filtering algorithms that real social media companies use, are corporate secrets, but we know that there are three main principles of content curation: popularity, semantic and collaborative filtering \cite{Perra2019}. Popularity filtering refers to the practice of promoting content that is popular across the platform; semantic filtering means that post similar to previous consumed posts are recommended and collaborative filtering suggests posts that are similar to the ones our friends consume \cite{Perra2019}. In Chapter \ref{chap3}, Section \ref{filter} a detailed description of the filtering algorithms used in  this thesis will be given.

@@

Another important tool in the study of opinion dynamics/formations in large groups of people is the use of statistical physics. This branch of physics offers a framework that relates microscopic properties of atoms, molecules,... to macroscopic observed behavior and has a wide field of applications inside and outside the world of physics. The observation that large number of people display collective, `macroscopic' behavior begged for the use of the concepts and insights developed in statistical physics \cite{Sirbu2016}. The application of the theory of statistical physics on social phenomena is referred to as social physics or sociophysics. The `microscopic' constituents are now individual humans who interact with a limited number of other individuals and in that way form complex, `macroscopic' groups such as human societies. These `macroscopic' groups display stunning regularities, transitions from disorder to order, the emergence of consensus, universality, etc \cite{Buchanan2007}. The statistical physics approach of these social systems tries to explain the found regularities at large scale as collective effects of the interactions among these individuals \cite{Sirbu2016}.\\
\newline
The statistical approach of social phenomena has one big difficulty compared to the description of gases and other physical phenomena. In regular physics, the microscopic constituents are `simple' atoms or molecules, whose dynamics and interactions with each other are often well known and can be put in relatively simple laws with only a few parameters \cite{Castellano2009}. Humans, on the other hand, are far from simple entities and the dynamics of a single individual are far from known \cite{Castellano2009}. Even if the dynamics/behavior of individuals would be well know, then the interactions between them would still be complicated and it would be impossible to describe them with simple laws and only a few parameters \cite{Castellano2009}. Hence, it may seem that extracting useful and qualitative information out of social models is a nearly impossible and almost hopeless task \cite{Castellano2009}. The use of statistical physics may, however, also offer a way out. In most situations the qualitative (and sometimes even quantitative) properties of large scale phenomena do not depend on the microscopic details of the process \cite{Castellano2009}. Instead, the relevant aspects for the global behavior are often higher level features such as symmetries, conservation laws, dimensionality, etc \cite{Castellano2009}. This introduces a concept of universality. With this universality in mind, one can proceed to build models of human/social behavior where only the simplest and most important properties of single individuals and their interactions are taken into account \cite{Castellano2009}. An important factor herein is the comparison with real/empirical data to investigate whether or not the trends seen in real life are compatible with the outputs generated by the models \cite{Castellano2009}. Note that this interplay between theory, model, experiment and real-life data is not only an important factor in social physics, but also in regular physical research and science in general.

@@

As said before, this thesis will try to investigate the interplay between the social network structure and the individual resistance to change in the evolution of two competing opinions. On one hand, we want to determine the influence of different network structures on the formation and evolution of opinions, whereas, on the other hand, we would like to investigate the effect of stubborn actors in the network. Since this thesis deals with opinion dynamics on social media, the effect of filtering algorithms must also be included.\\ 
\newline
In one of their previous papers (\textit{``Modelling opinion dynamics in the age of algorithmic personalization''}, see reference \cite{Perra2019}) Perra and Rocha investigated both the effect of different filtering algorithms and different network topologies on the formation and evolution of two competing opinions. They also considered the effect of nudging (meaning that one opinion is pushed to all agents in the network). Their main findings were that the algorithmic filtering could not break the status quo when the prevalence of opinions was equally distributed in the population and that topological correlations (such as high clustering coefficient) could result in the formation of echo chambers \cite{Perra2019}. On the other hand, a more heterogeneous contact pattern could hamper this formation of echo chambers \cite{Perra2019}. If the two opinions were, initially, not equally distributed, one of the filtering algorithms (namely semantic filtering) caused a further increase in the predominant opinion \cite{Perra2019}. In the case of nudging, they found that the opinions in the population moved to the nudged opinion relatively fast, even in the case of small nudging \cite{Perra2019}. They, however, did not investigate the effect of stubborn actors in the network, nor did they investigate networks with a community structure. This thesis will build on their model and incorporate these new features. We will compare networks that have a community structure and relatively low clustering to networks without community structure, but with a high clustering coefficient (a detailed description of concepts like clustering coefficient and community structure will be given in Chapter \ref{chap2}, Section \ref{netDef}). We want to investigate whether networks with community structure are able to form echo chambers, such as one can observe in networks with a high clustering coefficient and, if so, if one can, for example, observe convergence to one opinion inside the communities. We will also pay some attention to networks that have both a high clustering coefficient and a community structure. Can we observe echo chambers in those kinds of networks? If yes, can we determine which property might be the main driver for the formation of echo chambers: high clustering or community structure? \\
\newline
On the other hand, we want to determine the impact of stubborn actors on the formation and evolution of opinions in the network. Some interesting questions that may come to mind are: Is there an interplay between network structure and stubbornness? What about the possible interplay between the filtering algorithms and the introduction of stubborn actors? Does the distribution of stubborn actors play a role (eg. randomly distributed versus clustered together inside communities)? \\
% partly rewrite this according to results? Convergence to one opinion in communities --> discussed? Clustered stubborn people --> discussed? Include clustered opinions? Include real-world networks?

@@

Networks or graphs are conceptually very simple and flexible objects that are made of a set of nodes/vertices $V$ and a set of edges/links $E$, where the elements of $E$ determine connections between elements of $V$ \cite{Costa2018}
\begin{equation}
	E \subseteq \{\{x, y\}| x, y \in V \}\ .
\end{equation}
If self-loops are not allowed, we need the following condition on the elements of $E$
\begin{equation}
	E \subseteq \{\{x, y\}| x, y \in V \text{and\ } x \neq y \} \ .
\end{equation}
A graph with no self-loops is a simple graph. The nodes $x$ and $y$ of an edge $\{x, y\}$ are called the endpoints of the edge. It is possible that nodes are not joined by any edge, such nodes are called isolated or disconnected. If two or more edges have the same pair of endpoints, the graph is called a multi-graph. This thesis is not concerned with multi-graphs, nor are self-loops allowed. Graphs in which the edges have an orientation are called directed graphs \cite{Costa2018}. If the edges have weights $w_{ij}$, one speaks of a weighted graph.\\
\newline
A graph can be represented in different ways. One of the most common ways is by use of the adjacency matrix.

@@

Several very useful measurements quantify the notion of length and distance in a network/graph. We start with the definition of a path: a path is a sequence of distinct nodes, such that adjacent nodes in the sequence are adjacent nodes in the network \cite{Goddard2010}. The length of a path is then defined as the number of edges in the path \cite{Goddard2010}. A geodesic path, often called the shortest path, between two nodes is the path between two nodes $i$ and $j$ with minimum length (note that the geodesic path is not necessarily unique) \cite{F.Costa2007}. The length of the geodesic path between the two nodes $i$ and $j$ is called the geodesic distance or shortest distance $d_{ij}$ \cite{F.Costa2007}. Geodesic distance is often just called distance \cite{Goddard2010}. The notion of distance enables to define several network measurements such as average distance, diameter and radius.\\  
\newline
\textbf{Average distance}\\
\newline
The average distance is the mean value of the geodesic distance $d_{ij}$ \cite{F.Costa2007}
\begin{equation}\label{avdist}
	l = \frac{1}{N(N-1)} \sum_{i \neq j} d_{ij} \ .
\end{equation}
This definition, however, diverges if there are unconnected nodes in the network (unconnected nodes have, by definition, a distance equal to infinity) \cite{F.Costa2007}. This problem could be avoided by only including connected pairs in the sum, but this introduces a distortion for networks with a lot of unconnected pairs of nodes. These networks will then have a small average distance, which is only expected for networks with lots of connections \cite{F.Costa2007}. Instead, another definition of average distance is used: the global efficiency \cite{F.Costa2007}
\begin{equation}
	E = \frac{1}{N(N-1)} \sum_{i \neq j} \frac{1}{d_{ij}} \ .
\end{equation}
This measurement quantifies the efficiency of the network in sending information between nodes \cite{F.Costa2007}. This assumes that the efficiency of sending information between two nodes is reciprocal to their distance. The harmonic mean of the geodesic distances is now defined as the reciprocal of the global efficiency \cite{F.Costa2007}
\begin{equation}
 	h = \frac{1}{E} \ .
\end{equation}
\newline
\textbf{Diameter and radius}\\
\newline
The diameter $diam(G)$ of a network $G$ is the maximum distance between any pair of nodes \cite{Goddard2010}. The eccentricity of a node is the maximum distance from that node to any other node. The radius $rad(G)$ is the minimum eccentricity among all nodes of $G$. Note that the diameter is the maximum eccentricity among all nodes in $G$ \cite{Goddard2010}.
\newpage
\noindent
\subsection{Clustering coefficient and transitivity}

@@

where $n_i$ is the number of edges that actually exist between the nodes in the neighborhood of $i$ and $k_i(k_i-1)/2$ is the maximum number of edges that could exist between them (note that this expression is only valid for an undirected graph, since for a directed graph $e_{ij} \neq e_{ji}$ and we have $k_i(k_i-1)$ possible edges between the neighbors of node i). % should give more mathimatical expression for n_i?

@@

where $N$ denotes, once again, the number of nodes in the network. The clustering coefficient is, beside a measure for the connectivity in the network, also linked to the robustness, or resilience against random damage,  of the network \cite{Heer2020}\cite{Iyer2013} \cite{Li2017}. \\

@@

The derivation of this formula and thus of Eq. (\ref{Qc}) can be found in Appendix \ref{modul}, Section \ref{simplemod}. The modularity $Q$ can have values in the interval $\interval{-1}{1}$ and has the following properties \cite{Albert2016}
	\item The higher the value of $Q$ is for a certain partition, the better is the corresponding community structure. This idea is used in community detection algorithms, where modularity is maximized in order to find the optimal community structure.
	\item A modularity $Q$ equal to zero corresponds to the situation where all nodes are put in the same community, this is if the whole network is put into a single community. Then the two terms in Eq. (\ref{mod}) become equal and $Q$ vanishes.
	\item If every node is put into a separate community, the first term in Eq. (\ref{mod}) vanishes and we are left with $n_c$ negative terms. Hence, this situation corresponds to a negative modularity.

@@

In 1959 the two mathematicians Paul Erd\H{o}s and Alfr\'{e}d R\'{e}nyi proposed a model to generate simple random networks \cite{F.Costa2007}. Independently of them, Solomonoff and Rapoport already proposed the model in 1951. The model constructs random networks with $N$ nodes in the following way \cite{Albert2002}
\begin{enumerate}
	\item Start with $N$ disconnected nodes,
	\item For each pair of nodes, connect them with a predefined probability $p$.
\end{enumerate} 

@@

This equation can be simplified to $\left<k\right> = (N-1)p$, and for $N \gg 1$ this approximately becomes $\left<k\right> = Np$. The derivation of this result can be found in Appendix \ref{ER}, Section \ref{degree}.\\

@@

The stochastic block model in its most simple form is defined by the following parameters \cite{Clauset2017}
	\item The number of nodes $N$, 
	\item A partition of the node set $\{1, 2, ..., N\}$ into $k$ disjoint subsets $C_1, ..., C_k$; these groups are the communities,

>>>>

\renewcommand{\abstractname}{Abstract}
\selectlanguage{english}
\begin{abstract}
\addcontentsline{toc}{chapter}{Abstract}
The past decade, the rise of social media has become undeniable. More and more people are getting engaged on some type of social media platform and hence, it is probably not wrong to say that nowadays most of us have, beside our real life, a virtual life on social media. This rapidly increased engagement on social media undoubtedly has many advantages. However, every upside has a downside and social media is no different. This thesis deals with the impact of social media on the formation and evolution of opinions in society.\\ 
Social media allows for the spread of information at a faster pace and at an unprecedented scale. However, people are still bound by time- and cognitive constraints. In order to handle the information overflow, social media companies design filtering algorithms that can fine-tune and individualize the information someone is exposed to. This filtering and individualizing of information may have some potentially dangerous downsides in the sense that people might get trapped in ``echo chambers'' that confirm their own beliefs. It is important to know the effects of such algorithmic personalization on the appearance and formation of phenomena such as polarization, spread of extremism, formation of echo chambers, etc.\\
People are, however, more than sheep that thoughtlessly follow their shepherd (contrary to what some books might lead us to believe). They are capable of critical thinking and may be hesitant to change their current opinion. This aspect of potential stubbornness should not be ignored.\\
In this thesis toy models are designed to investigate the interplay between network structure, filtering algorithms and individual resistance to change in the evolution of two competing opinions.\\
We will investigate whether network properties such as community structure and clustering coefficient may enhance or hamper the formation of echo chambers. We will use some basic filtering algorithms to determine their possible effects on the appearance of echo chambers and we will investigate the impact of individual stubbornness. \\
Beside that, we will also pay some attention to the possible ways of distributing the opinions and how this may affect the opinion dynamics in the system. \\
Finally, we will test the model on real-world networks. Some question that may arise are: Do the results agree with those obtained from the theoretical network models? What are the similarities or differences in the results? What are the possible underlying reason/structures for these similarities/differences?\\

@@

The past decade social media has become ever more important in our daily lives. Nowadays most of us have, beside a real-world life, a virtual life on social media. This increased engagement on social media may have a huge impact on the formation of opinions, not only on the individual scale, but also on a broader, societal scale. \\
\newline
It is known that individuals not only form their opinions through self-reflection, but also through interactions with other people and with their surroundings \cite{Perra2019}. The broad range of interactions people undergo on social media with people from all over the world can thus not be underestimated in the process of opinion formation. The understanding of the role of social media on the emergence of polarization and extremism in society is of uttermost importance.\\
\newline
It is also known that some people change their opinions more easily than others, that is, once in a while you encounter somebody that is really stubborn and persistent towards its own opinion. It is important to understand the effects of individual resistance on the formation and evolution of opinions. When designing models of opinion dynamics on social networks, this individual stubbornness should not be ignored.\\
\newline
One important feature of opinion dynamics on social media is the use of algorithmic personalization. Social media allow for simultaneous or asynchronous interactions between people without geographical constraints and thus allow for the spread of information at a faster pace and at an unprecedented scale \cite{Perra2019}. People are, however, still bound by temporal and cognitive constraints. In order to assure a more pleasant/convenient experience on their platforms, social media companies use algorithms to order/filter the posts on an individual's time-line according to what might be relevant to that individual \cite{Perra2019}. It is of great importance to know the possible effects of these filtering algorithms on the evolution of opinions in the population.\\  
\newline
Opinion dynamics models have two important layers. On one hand, we introduce social networks to describe the underlying structure of social interactions; on the other hand, we need appropriate opinion dynamics/formation models to reproduce the opinion dynamics and opinion formation processes found in real life.\\
\newline
Section \ref{complNet} gives an introduction to the first layer, namely that of complex networks. The second layer, the opinion dynamics layer, is introduced in Section \ref{Opinion}. Section \ref{statPhys} introduces concepts such as statistical and social physics. Finally, in Section \ref{goal}, the hypotheses and goals of this thesis are summarized and discussed.

@@

Some other important concepts in network theory are the degree distribution, clustering/community structure, connectivity, etc. These concepts will be explained in depth in Chapter \ref{chap2}, Section \ref{netDef}, but let us already anticipate a bit on the case of social networks. It is found that many real-life social systems have a power-law degree distribution \cite{Muchnik2013}. This heterogeneous or scale-free degree distribution represents one of the three general properties of social networks. The other two are short distances and high clustering. These two are also referred to as ``small world phenomenon'' (note that in the literature the term ``small world'' sometimes only refers to the property of short distances, however, in this thesis, we will use it to indicate both a high clustering and short distances) \cite{Muchnik2013}. Ideally, our theoretical network should exhibit these three properties. Some theoretical networks, that are thoroughly studied, do not possess all of them. However, they might still be used, because of their simplicity and ability to produce analytic results. When using these models one must always bear in mind their limitations to reproduce some properties encountered in real social systems. \\

@@

As stated before, the formation of an individual's opinion is the result of the interplay of many factors such as self-reflection, peer pressure, the individual's personality (eg. stubbornness), the information someone is exposed to, etc.  Opinion dynamics models should try to capture these complex processes in a simplified way. Several models have been developed ranging from simple binary models to more complex, continuous approaches \cite{Sirbu2016}.\\
\newline
The basic idea of all opinion dynamics models is that the nodes or agents in a social network have a variable that represents their opinion and that is updated according to some predefined rules. These models are a simplification of real-world opinion dynamics. It is however shown that they display a lot of aspects of real opinion formation such as agreement, transitions between order (consensus) and disorder (fragmentation), polarization, formation of echo chambers (clusters of people with the same opinion), etc \cite{Sirbu2016}. \\
\newline
The opinion of the agents in the model can be either discrete or continuous. This thesis deals with models where each agent can have one of the two opinions, A or B, and thus only deals with the case of discrete opinions. This might come across as a huge simplification of the real-world complexity of opinion formation, but also in real life people often have to choose between two competing opinions (eg. republicans or democrats, cat or dog, renting or buying a house,...). \\
\newline
The rules that determine how an agent updates his or her opinion can include many different aspects. They can, for example, be as simple as a majority rule (i.e. the majority model: if the majority of your neighbors have a certain opinion, you adopt that opinion as well) or can include a more probabilistic way of updating. A small overview of possible opinion dynamics models is given in Chapter \ref{chap3}, Section \ref{Bin} and Section \ref{Con}. \\
\newline
One can also include a resistance parameter that determines the hesitation of an agent to change to a new opinion and/or a parameter that determines the influence of an agent on others. Furthermore, the agents can have a parameter that determines whether they are active or not. The introduction of stubborn actors will be discussed in Chapter \ref{chap3}, Section \ref{stubb}; the activation mechanism of nodes is discussed in Chapter \ref{chap3}, Section \ref{actMech}.\\
\newline
Since this thesis deals with the particular case of opinion dynamics on on-line social platforms, it is important to define/set up different filtering algorithms. The filtering algorithms that real social media companies use, are corporate secrets, but we know that there are three main principles of content curation: popularity, semantic and collaborative filtering \cite{Perra2019}. Popularity filtering refers to the practice of promoting content that is popular across the platform; semantic filtering means that post similar to previous consumed posts are recommended and collaborative filtering suggests posts that are similar to the ones our friends consume \cite{Perra2019}. In Chapter \ref{chap3}, Section \ref{filter} a detailed description of the filtering algorithms used in  this thesis will be given.

@@

Another important tool in the study of opinion dynamics/formations in large groups of people is the use of statistical physics. This branch of physics offers a framework that relates microscopic properties of atoms, molecules,... to macroscopic observed behavior and has a wide field of applications inside and outside the world of physics. The observation that large number of people display collective, `macroscopic' behavior begged for the use of the concepts and insights developed in statistical physics \cite{Sirbu2016}. The application of the theory of statistical physics on social phenomena is referred to as social physics or sociophysics. The `microscopic' constituents are now individual humans who interact with a limited number of other individuals and in that way form complex, `macroscopic' groups such as human societies. These `macroscopic' groups display stunning regularities, transitions from disorder to order, the emergence of consensus, universality, etc \cite{Buchanan2007}. The statistical physics approach of these social systems tries to explain the found regularities at large scale as collective effects of the interactions among these individuals \cite{Sirbu2016}.\\
\newline
The statistical approach of social phenomena has one big difficulty compared to the description of gases and other physical phenomena. In regular physics, the microscopic constituents are `simple' atoms or molecules, whose dynamics and interactions with each other are often well known and can be put in relatively simple laws with only a few parameters \cite{Castellano2009}. Humans, on the other hand, are far from simple entities and the dynamics of a single individual are far from known \cite{Castellano2009}. Even if the dynamics/behavior of individuals would be well know, then the interactions between them would still be complicated and it would be impossible to describe them with simple laws and only a few parameters \cite{Castellano2009}. Hence, it may seem that extracting useful and qualitative information out of social models is a nearly impossible and almost hopeless task \cite{Castellano2009}. The use of statistical physics may, however, also offer a way out. In most situations the qualitative (and sometimes even quantitative) properties of large scale phenomena do not depend on the microscopic details of the process \cite{Castellano2009}. Instead, the relevant aspects for the global behavior are often higher level features such as symmetries, conservation laws, dimensionality, etc \cite{Castellano2009}. This introduces a concept of universality. With this universality in mind, one can proceed to build models of human/social behavior where only the simplest and most important properties of single individuals and their interactions are taken into account \cite{Castellano2009}. An important factor herein is the comparison with real/empirical data to investigate whether or not the trends seen in real life are compatible with the outputs generated by the models \cite{Castellano2009}. This interplay between theory, model, experiment and real-life data is not only an important factor in social physics, but also in regular physical research and science in general.

@@

This thesis will try to investigate the interplay between the social network structure and the individual resistance to change in the evolution of two competing opinions. On one hand, we want to determine the influence of different network structures on the formation and evolution of opinions, whereas, on the other hand, we would like to investigate the effect of stubborn actors in the network. Since this thesis deals with opinion dynamics on social media, the effect of filtering algorithms must also be included.\\ 
\newline
In one of their previous papers (\textit{``Modelling opinion dynamics in the age of algorithmic personalization''}, \cite{Perra2019}) Perra and Rocha investigated both the effect of different filtering algorithms and different network topologies on the formation and evolution of two competing opinions. They also considered the effect of nudging (meaning that one opinion is pushed to all agents in the network). Their main findings were that the algorithmic filtering could not break the status quo when the prevalence of opinions was equally distributed in the population and that topological correlations (such as high clustering coefficient) could result in the formation of echo chambers \cite{Perra2019}. On the other hand, a more heterogeneous contact pattern hampers this formation of echo chambers \cite{Perra2019}. If the two opinions were, initially, not equally distributed, one of the filtering algorithms (namely semantic filtering) caused a further increase in the predominant opinion \cite{Perra2019}. In the case of nudging, they found that the opinions in the population moved to the nudged opinion relatively fast, even in the case of small nudging \cite{Perra2019}. They, however, did not investigate the effect of stubborn actors in the network, nor did they investigate networks with a community structure. This thesis will build on their model and incorporate these new features. We will compare networks that have a community structure and relatively low clustering to networks without community structure, but with a high clustering coefficient (a detailed description of concepts like clustering coefficient and community structure will be given in Chapter \ref{chap2}, Section \ref{netDef}). We want to investigate whether networks with community structure are able to form echo chambers, such as one can observe in networks with a high clustering coefficient. We will also pay some attention to networks that have both a high clustering coefficient and community structure. Can we observe echo chambers in those kinds of networks? If yes, can we determine which property might be the main driver for the formation of echo chambers: high clustering or community structure? \\
\newline
On the other hand, we want to determine the impact of stubborn actors on the formation and evolution of opinions in the network. Some interesting questions that may come to mind are: Is there an interplay between network structure and stubbornness? What about the possible interplay between the filtering algorithms and the introduction of stubborn actors?\\
% partly rewrite this according to results? Convergence to one opinion in communities --> discussed? Clustered stubborn people --> discussed? Include clustered opinions? Include real-world networks?

@@

Networks or graphs are conceptually simple and flexible objects that are made of a set of nodes/vertices $V$ and a set of edges/links $E$, where the elements of $E$ determine connections between elements of $V$ \cite{Costa2018}
\begin{equation}
	E \subseteq \{\{x, y\}| x, y \in V \}\ .
\end{equation}
If self-loops are not allowed, we need the following condition on the elements of $E$
\begin{equation}
	E \subseteq \{\{x, y\}| x, y \in V \text{and\ } x \neq y \} \ .
\end{equation}
A graph with no self-loops is a simple graph. The nodes $x$ and $y$ of an edge $\{x, y\}$ are called the endpoints of the edge. It is possible that nodes are not joined by any edge, such nodes are called isolated or disconnected. If two or more edges have the same pair of endpoints, the graph is called a multi-graph. This thesis is not concerned with multi-graphs, nor are self-loops allowed. Graphs in which the edges have an orientation are called directed graphs \cite{Costa2018}. If the edges have weights $w_{ij}$, one speaks of a weighted graph.\\
\newline
A graph can be represented in different ways. One of the most common ways is by use of the adjacency matrix.

@@

Several useful measurements quantify the notion of length and distance in a network/graph. We start with the definition of a path: a path is a sequence of distinct nodes, such that adjacent nodes in the sequence are adjacent nodes in the network \cite{Goddard2010}. The length of a path is then defined as the number of edges in the path \cite{Goddard2010}. A geodesic path, often called the shortest path, between two nodes is the path between two nodes $i$ and $j$ with minimum length (the geodesic path is not necessarily unique) \cite{F.Costa2007}. The length of the geodesic path between the two nodes $i$ and $j$ is called the geodesic distance or shortest distance $d_{ij}$ \cite{F.Costa2007}. Geodesic distance is often just called distance \cite{Goddard2010}. The notion of distance enables to define several global network measurements such as average distance, diameter and radius.\\  
\newline
\textbf{Average distance}\\
\newline
The average distance is the mean value of the geodesic distance $d_{ij}$ \cite{F.Costa2007}
\begin{equation}\label{avdist}
	l = \frac{1}{N(N-1)} \sum_{i \neq j} d_{ij} \ .
\end{equation}
This definition, however, diverges if there are unconnected nodes in the network (unconnected nodes have, by definition, a distance equal to infinity) \cite{F.Costa2007}. This problem could be avoided by only including connected pairs in the sum, but this introduces a distortion for networks with a lot of unconnected pairs of nodes. These networks will then have a small average distance, which is only expected for networks with lots of connections \cite{F.Costa2007}. Instead, another definition of average distance is used: the global efficiency $E$ \cite{F.Costa2007}
\begin{equation}
	E = \frac{1}{N(N-1)} \sum_{i \neq j} \frac{1}{d_{ij}} \ .
\end{equation}
This measurement quantifies the efficiency of the network in sending information between nodes \cite{F.Costa2007}. This assumes that the efficiency of sending information between two nodes is reciprocal to their distance. The harmonic mean of the geodesic distances is now defined as the reciprocal of the global efficiency \cite{F.Costa2007}
\begin{equation}
 	h = \frac{1}{E} \ .
\end{equation}
\newline
\textbf{Diameter and radius}\\
\newline
The diameter $diam(G)$ of a network $G$ is the maximum distance between any pair of nodes \cite{Goddard2010}. The eccentricity of a node is the maximum distance from that node to any other node. The radius $rad(G)$ is the minimum eccentricity among all nodes of $G$. Note that the diameter is the maximum eccentricity among all nodes in $G$ \cite{Goddard2010}.
\newpage
\noindent
\subsection{Clustering coefficient and transitivity}

@@

where $n_i$ is the number of edges that actually exist between the nodes in the neighborhood of $i$ and $k_i(k_i-1)/2$ is the maximum number of edges that could exist between them (this expression is only valid for an undirected graph, since for a directed graph $e_{ij} \neq e_{ji}$ and we have $k_i(k_i-1)$ possible edges between the neighbors of node i). % should give more mathimatical expression for n_i?

@@

where $N$ denotes the number of nodes in the network. The clustering coefficient is, beside a measure for the connectivity in the network, also linked to the robustness, or resilience against random damage,  of the network \cite{Heer2020}\cite{Iyer2013}\cite{Li2017}. \\

@@

The derivation of this formula, and thus of Eq. (\ref{Qc}), can be found in Appendix \ref{modul}, Section \ref{simplemod}. The modularity $Q$ can have values in the interval $\interval{-1}{1}$ and has the following properties \cite{Albert2016}: (i) The higher the value of $Q$ for a certain partition, the better the corresponding community structure. This idea is used in community detection algorithms, where modularity is maximized in order to find the optimal community structure; (ii) A modularity $Q$ equal to zero corresponds to the situation where all nodes are put in the same community, this is, if the whole network is put into a single community. Then the two terms in Eq. (\ref{mod}) become equal and $Q$ vanishes; (iii) If every node is put into a separate community, the first term in Eq. (\ref{mod}) vanishes and we are left with $n_c$ negative terms. Hence, this situation corresponds to a negative modularity.\\

@@

In 1959 the two mathematicians Paul Erd\H{o}s and Alfr\'{e}d R\'{e}nyi proposed a model to generate simple random networks \cite{F.Costa2007}. Independently of them, Solomonoff and Rapoport already proposed the model in 1951. The model constructs random networks with $N$ nodes in the following way \cite{Albert2002}: (i) Start with $N$ disconnected nodes; (ii) For each pair of nodes, connect them with a predefined probability $p$.\\

@@

This equation can be simplified to $\left<k\right> = (N-1)p$, and for $N \gg 1$ this approximately becomes $\left<k\right> \approx Np$. The derivation of this result can be found in Appendix \ref{ER}, Section \ref{degree}.\\

@@

The stochastic block model in its most simple form is defined by the following parameters \cite{Clauset2017}: (i) The number of nodes $N$; (ii) A partition of the node set $\{1, 2, ..., N\}$ into $k$ disjoint subsets $C_1, ..., C_k$; these groups are the communities; (iii) A symmetric $k\times k$ matrix $M$; the diagonal elements $M_{ii}$, $i = 1, ..., k$, represent the edge probabilities inside the community $C_i$ and the off-diagonal elements $M_{ij}$, $i, j = 1, ..., k$ with $i \neq j$, give the edge probabilities between nodes in community $C_i$ and nodes in community $C_j$.\\
